---
title: "ST 502 Final"
author: "Katelyn Settlemyre, Julia Farrell"
date: "2022-12-04"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#include libraries needed here

```

## Introduction

_What is the general idea of testing for homogeneity? Perhaps also explain the purpose and outline of this project._

This report will explore the Chi-Square test for homogeneity in detail. We will derive the likelihood ratio test stastic and explain the Pearson Chi-Square statistic associated with this test that is used to determine the rejection criterion. Once the theory is well established, we then conduct a simulation. The goal is to determine how well the asymptotic rejection region performs at controlling the alpha level of the Pearson Chi-Square test and to determine the power of the asymptotic test when comparing certain alternative hypotheses.


## Data Example

_Use hospital data given to conduct a_ $\chi^2$ _test for homogeneity._
```{r}
#create and print matrix of hospital data
rows <- rbind(a=c(41, 27, 51), b=c(36,3,40), c=c(169, 106, 109))
(hospDat <- matrix(data=rows, nrow = 3, ncol = 3, 
                   dimnames = list(c("A", "B", "C"), 
                                   c("Surgical Site Infections", 
                                     "Pneumonia Infections", 
                                     "Bloodstream Infections"))))
```


## Deriving the Likelihood Ratio Test

_Derive the LRT test for homogeneity; some work already done on page 151 of our notes. Pages 129, 132, 137, 139 may also be useful?_


## Simulation

_Simulate test with data. Potentially make use of code already given in notes. Two multinomial case only, with 3 categories in each multinomial. Plot summaries of simulations. Set seed for reproduction purposes._

Goal:

Process:

```{r}
#set seed for reproduction
set.seed(17)
```

